---
title: "Fact Sheet 4: Economics of Direct Seafood Sales"
format:
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
    fig-width: 7
    fig-height: 5
execute:
  cache: true
---

\vspace{-2cm}

The 2023 American Seafood Harvesters Marketing Practices Survey is the first-ever national assessment of the direct seafood sector. The survey aims to fill a gap in our understanding of the domestic seafood market in the US and bring greater focus to the socioeconomic contributions of seafood in the nation's local and regional food systems. This document describes economic aspects of direct seafood sales and is the fourth and final in a series of fact sheets that highlight findings from the survey

```{r libraries}
#| include: false
#| message: false
#| warning: false
library(here)
library(ggplot2)
library(forcats)
library(dplyr)
library(tidyr)
library(scales)
library(viridis)
library(janitor)
library(gt)
library(gtExtras)
library(RColorBrewer)
library(cowplot)
library(magick)
library(patchwork)
library(waffle)
#library(MASS)
library(ggridges)

```

```{r dataframes}
#| message: false
#| warning: false
#| echo: false
#| include: false
val_resp_data <- read.csv(here("output_files","valid_svy_resp_data.csv"), header = TRUE, na.strings = c("","NA"))
```

# Production & Marketing Expenses for Direct Seafood

```{r expenses}
#| include: false
#| echo: false
#| message: false
#| warning: false
expense_summary <- val_resp_data %>% 
  drop_na(X6.1.1) %>% 
  #group_by(NOAA_region) %>% 
  summarize(expense_median = median(X6.1.1),
            expense_mean = mean(X6.1.1),
            expense_max = max(X6.1.1),
            expense_min = min(X6.1.1))

ggplot(val_resp_data %>% drop_na(X6.1.1)) +
  geom_boxplot(aes(y=(X6.1.1+10))) +
   scale_y_continuous(limits = c(0,100000)) + coord_flip()


val_resp_data %>% 
  drop_na(X6.1.1) %>% 
  #group_by(NOAA_region) %>% 
  ggplot(aes(y = NOAA_region, x = (X6.1.1+1), group = NOAA_region, fill = NOAA_region)) +
  #ggplot(aes(x = (X6.1.1+1), y= ExternalReference, group = ExternalReference)) +
  geom_violin(width = 1) +
  #geom_density_ridges(jittered_points = TRUE, quantile_lines = TRUE, scale = 0.7, alpha = 0.5, quantiles = 2,
   # vline_size = 1, vline_color = "red", 
    #point_size = 0.5, point_alpha = 0.6,
    #position = position_raincloud(adjust_vlines = FALSE)) +
  stat_summary(fun = mean, geom="point", shape=20, size=4, color="red", fill="red") +
  #scale_x_log10() +
  scale_x_continuous(expand = c(0,0), limits = c(0,100000)) +
  #scale_y_discrete(labels = c('\u2265 400 miles', '100 - 400 miles', '\u2264 100 miles')) +
  theme_ridges() + theme(legend.position = "none",
                          plot.margin = margin(0.5, 1, 0.5, 0.5, "cm"))

```

```{r expense-pct}
#| include: false
#| echo: false
#| message: false
#| warning: false

expense_pct <- val_resp_data %>% 
  dplyr::select(ExternalReference, NOAA_region, X6.11_1, X6.11_2, X6.11_3, X6.11_4, X6.11_5, X6.11_6, X6.11_7, X6.11_8,X6.11_9) %>% 
  mutate(NOAA_region = factor(NOAA_region, 
                              levels = c("North Pacific (Alaska)", "Pacific", "Western Pacific (Hawaii)", "New England", "Mid-Atlantic", "South Atlantic", "Gulf of Mexico", "Great Lakes", "Other"),
                              labels = c("North Pacific (Alaska)", "Pacific", "Western Pacific (Hawaii)", "New England", "Mid-Atlantic", "South Atlantic", "Gulf of Mexico", "Great Lakes", "Other"))) %>% 
  mutate_at(c('X6.11_1', 'X6.11_2', 'X6.11_3', 'X6.11_4', 'X6.11_5', 'X6.11_6', 'X6.11_7', 'X6.11_8','X6.11_9'), as.numeric) %>% 
  rowwise() %>% 
  mutate(Total = sum(c_across(X6.11_1:X6.11_9), na.rm = T)) %>% 
  filter(Total == '100') %>% 
  rename('crew' = "X6.11_1",
         'harvest_exp' = "X6.11_2",
         'fuel' = "X6.11_3",
         'processing' = "X6.11_4",
         'mkt_labor' = "X6.11_5",
         'distribution' = "X6.11_6",
         'marketing' = "X6.11_7",
         'sale_equip' = "X6.11_8",
         'other' = "X6.11_9") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  ungroup() %>% 
  mutate(NOAA_region = fct_collapse(NOAA_region,
                                    "Great Lakes" = c("Other", "Great Lakes"))) %>% 
  group_by(NOAA_region) %>% 
  summarize(crew = mean(crew),
            harvest_exp = mean(harvest_exp),
            fuel = mean(fuel),
            processing = mean(processing), 
            mkt_labor = mean(mkt_labor),
            distribution = mean(distribution),
            marketing = mean(marketing), 
            sale_equip = mean(sale_equip),
            other = mean(other)) %>% 
  pivot_longer(crew:other, 
               names_to = "expenses", values_to = "pct") %>% 
  mutate(expenses = factor(expenses,
                          levels = c("crew", "harvest_exp", "fuel", "processing", "mkt_labor", "distribution", "marketing", "sale_equip", "other")
                          #,labels = c("Source Identified\n Distributors", "Institution", "Foodservices", "Retail", "Consumer")
                          )) %>% 
  ungroup() %>% 
  mutate(pct = round(pct,0))

us_exp_summary <- expense_pct %>% group_by(expenses) %>% 
  summarise(NOAA_region = 'United States',
            pct = mean(pct)) %>% 
  mutate(pct_round = as.numeric(c(19, 28, 25, 6, 1, 6, 2, 7, 6))) %>% 
  mutate(labels = factor(expenses,
                          levels = c("harvest_exp", "fuel", "crew","sale_equip","mkt_labor","processing", "distribution",  "marketing",  "other"), labels = c("Seafood harvesting expenses (permits, insurance, vessels, gear, etc.)", "Fuel for seafood harvesting", "Hired crew on seafood harvesting operations", "Seafood sales equipment (boxes, totes, scales, tables, etc.)",  "Hired labor for seafood sales", "Seafood processing costs",  "Seafood transportation & distribution costs",  "Seafood marketing expenses (advertising, website, etc.)", "Other expenses (market fees, licenses, insurance, etc.)")))
  
exp_waff <- us_exp_summary %>% dplyr::select(labels, pct_round) %>% arrange(labels)
  

```

```{r exp-waffle, fig.height=3, fig.width=7.5}
#| echo: false
#| message: false
#| fig-align: center
#| include: true
waffle(exp_waff,
       flip = TRUE,
       reverse = TRUE,
       size = 0.5,
       equal = TRUE, legend_pos = "left") + 
  scale_fill_brewer(name = "Proportional Expenses", palette = "Set1") +
  theme_enhance_waffle() +
  theme(legend.text = element_text(size=10))


ggsave(here("output_figs", "expense_waffle.png"), width = 7.5, height = 3, units = "in", dpi = 300, bg = "white")
```

Direct Seafood Marketing has expenses associated not only with seafood marketing but also with seafood harvesting, processing, and distribution. We heard from nearly **500** direct seafood marketers about their seafood marketing associated expenses in 2022.

The reported expenses ranged widely across the US, from as little as **\$200** to an average of **\$600,000**. Accounting for this variability, the median expenses were **\$30,000**. Nearly three quarters of the reported expenses were associated with seafood harvesting, including equipment costs, fuel, and crew salaries. On the other hand, seafood marketing related expenses were proportionately lower and divided nearly equally across processing, distribution, marketing equipment, and other expenses.

\newpage

# Income from Direct Seafood Sales

Across the US, direct seafood businesses reported varying incomes from direct seafood sales in 2022. A third of businesses reported incomes upto **\$10,000**, while **40%** of businesses reported incomes ranging between **\$10,000** and **\$100,000**. A little less than a third reported incomes greater than **\$100,000**.

```{r income}
#| include: false
#| echo: false
#| message: false
#| warning: false
income_summary <- val_resp_data %>% 
  dplyr::select(ExternalReference, NOAA_region, state, X6.1, X6.3) %>% 
  mutate(X6.1 = factor(X6.1, 
                ordered = TRUE,
                levels = c("$1 - 999", "$1,000 - 2,499", "$2,500 - 4,999",
                           "$5,000 - 9,999", "$10,000 - 24,999", 
                           "$25,000 - 49,999", "$50,000 - 99,999",
                           "$100,000 - 249,999", "$250,000 - 499,999",
                           "$500,000 - 999,999", "$1,000,000 - 2,499,999",
                           "$2,500,000 - 4,999,999", "$5,000,000 and over"))) %>% 
  drop_na(X6.1) %>% 
  mutate(X6.1_clipped = fct_collapse(X6.1, 
                        "$1 - 999" = "$1 - 999",
                    "$1,000 - 9,999" = c("$1,000 - 2,499", "$2,500 - 4,999", "$5,000 - 9,999"),
                    "$10,000 - 49,999" = c("$10,000 - 24,999","$25,000 - 49,999"),
                    "$50,000 - 99,999" = "$50,000 - 99,999",
                    "$100,000 - 499,999" = c("$100,000 - 249,999", "$250,000 - 499,999"),
                    "$500,000 - 999,999" = "$500,000 - 999,999",
                    "$1,000,000 - 4,999,999" = c("$1,000,000 - 2,499,999", "$2,500,000 - 4,999,999"))) %>% 
  mutate(X6.1_trim = fct_collapse(X6.1, 
                       "$500,000 and over" = c("$500,000 - 999,999","$1,000,000 - 2,499,999", "$2,500,000 - 4,999,999","$5,000,000 and over"))) %>% 
  mutate(X6.3 = factor(X6.3, 
                ordered = TRUE,
                levels = c("$1 - 999", "$1,000 - 9,999", 
                           "$10,000 - 49,999", "$50,000 - 99,999",
                           "$100,000 - 499,999", "$500,000 - 999,999",
                           "$1,000,000 - 4,999,999", "$5,000,000 and over"))) %>% 
  drop_na(X6.3) %>% 
  rename(Gross = X6.3,
         Direct  = X6.1_clipped) %>% 
  pivot_longer(cols = c(Gross, Direct),
               names_to = "Income",
               values_to = "income_bracket")

income_summary %>% filter(Income == "Direct") %>% 
  group_by(X6.1_trim) %>% tally() %>% adorn_totals()
```

```{r income-fig, fig.width=7.5, fig.height=3}
#| echo: false
#| message: false
#| fig-align: center
#| include: true

income_summary %>% filter(Income == "Direct") %>% 
ggplot(aes(x=fct_rev(X6.1_trim))) +
  #geom_bar(aes(y = after_stat(count/sum(count))), fill = "#436AB3") + 
  geom_bar(aes(y = after_stat(count)), fill = "#436AB3") +
  scale_x_discrete(drop=FALSE) +
  scale_y_continuous(breaks = c(0,20,40,60,80), limits = c(0,81)) +
  coord_flip() + theme_cowplot() +
  xlab("") + ylab("Number of Direct Seafood Businesses") + 
  theme(panel.grid.major.x = element_line(color = "grey90"),
        axis.text.x = element_text(size=16),
        axis.title.y = element_text(size = 18))

ggsave(here("output_figs","dir-income-fig.png"), width = 7.5, height = 3, dpi = 300, units = "in", bg = "white")

```

# Percent Income from Direct Marketing Channels

Direct to consumer seafood sales contributed an average of **36%** to the income of seafood businesses, closely followed by Source-Identified Distributors with an average contribution of **32%**. Within regions of the US, the contributions to seafood business revenues varied by direct marketing channels. Either Direct to Consumer sales or Source-Identified Distributors contributed the most to regional direct seafood sales revenue.

```{r channel-pct}
#| include: false
#| echo: false
#| message: false
#| warning: false
channel_sales_pct <- val_resp_data %>% 
  dplyr::select(ExternalReference, NOAA_region, X6.2_1, X6.2_2, X6.2_3, X6.2_4, X6.2_5) %>% 
  mutate(NOAA_region = factor(NOAA_region, 
                              levels = c("North Pacific (Alaska)", "Pacific", "Western Pacific (Hawaii)", "New England", "Mid-Atlantic", "South Atlantic", "Gulf of Mexico", "Great Lakes", "Other"),
                              labels = c("North Pacific (Alaska)", "Pacific", "Western Pacific (Hawaii)", "New England", "Mid-Atlantic", "South Atlantic", "Gulf of Mexico", "Great Lakes", "Other"))) %>% 
  mutate_at(c('X6.2_1', 'X6.2_2', 'X6.2_3','X6.2_4','X6.2_5'), as.numeric) %>% 
  rowwise() %>% 
  mutate(Total = sum(c_across(X6.2_1:X6.2_5), na.rm = T)) %>% 
  filter(Total == '100') %>% 
  rename('dtc' = "X6.2_1",
         'retail' = "X6.2_2",
         'food' = "X6.2_3",
         'insti' = "X6.2_4",
         'sid' = "X6.2_5") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  ungroup() %>% 
  mutate(NOAA_region = fct_collapse(NOAA_region,
                                    "Great Lakes" = c("Other", "Great Lakes"))) %>% 
  group_by(NOAA_region) %>% 
  summarize(dtc = mean(dtc),
            retail = mean(retail),
            food = mean(food),
            insti = mean(insti), 
            sid = mean(sid)) %>% 
  pivot_longer(dtc:sid, 
               names_to = "channel", values_to = "pct") %>% 
  mutate(channel = factor(channel,
                          levels = c("sid", "insti", "food", "retail", "dtc"),
                          labels = c("Source Identified\n Distributors", "Institution", "Foodservices", "Retail", "Consumer"))) %>% 
  ungroup() %>% 
  mutate(pct = round(pct,0))

us_summary <- channel_sales_pct %>% group_by(channel) %>% 
  summarise(NOAA_region = 'United States',
            pct = round(mean(pct),0)) %>% 
  bind_rows(channel_sales_pct) %>% 
  mutate(NOAA_region = factor(NOAA_region, 
                              levels = c("United States", "North Pacific (Alaska)", "Pacific", "Western Pacific (Hawaii)", "New England", "Mid-Atlantic", "South Atlantic", "Gulf of Mexico", "Great Lakes")))
```

```{r channel-pct-fig, fig.width=7.5, fig.height=6}
#| echo: false
#| message: false
#| fig-align: center
#| include: true
ggplot(us_summary, aes(x= fct_rev(NOAA_region), 
                            y = pct, fill = fct_reorder(channel, pct, mean, na.rm = TRUE))) +
  geom_bar(position="stack", stat="identity", width = 4) +
  scale_fill_manual(breaks = c("Consumer", "Retail", "Foodservices", "Institution", "Source Identified\n Distributors"),
                    values = c("#C15227", "#062539", "#436AB3", "#99B0DB", "#FDEB4D")) +
  #geom_text(aes(label=paste0(sprintf("%1.0f", pct),"%")),
   #         position=position_stack(vjust=0.5)) +
  #scale_color_manual(values = c("black", "black", "white", "black", "black")) +
  scale_y_continuous(labels = label_percent(accuracy = NULL, scale =1)) +
  labs(x = "", y = "", fill = "") + 
  facet_wrap(NOAA_region~., scales = 'free_y', ncol = 1) +
  coord_flip() + theme_cowplot() +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        panel.spacing.y = unit(0.5, "lines"),
        strip.background = element_rect(color = "white", fill = "white"))

ggsave(here("output_figs", "channel_percents.png"), width = 7.5, height = 6, units = "in", dpi = 300, bg = "white")
```

COLLABORATING PARTNERS \| FUNDING PARTNERS

*\[INSERT UMAINE, LCN, USDA, NOAA FISHERIES LOGOS HERE\]*

For more information please contact Sahir Advani (sahir.advani\\\@maine.edu) or Josh Stoll (joshua.stoll\\\@maine.edu). DOI: 10.17605/OSF.IO/MBPJD
